Model for DES Type Ia Supernova Cosmology Analysis

### 2016-05-22 tex/toy.tex
pad up with more words to clarify.

There are top-level  sets of parameters $M$ that correspond to photometric calibration, cosmology,
host-galaxy properties, and transient properties.
There are two sets of intermediary parameters $M_I$ that correspond to luminosity $L$ and type $\tau$.

@@

We are interested in what the model predicts for the pdf of those observables
for the population of transients that enter the analysis.  
The probabilities for this ``analysis'' population are notated with the
subscript $S$, for example the likelihood is expressed as $\mathcal{L}_S(O| M, M_I)$.

The likelihood can be expressed in terms of probabilities of the observables
for all possible transients and observables accommodated by the model.
The probabilities for this ``underlying'' population are notated with no subscript,
so that the pdf of observables for this population is $p(O|M, M_I$).
The analysis population differs from the underlying population in that the transients of the former 
must, for the typed SN~Ia analysis, be discovered, pass photometric selection criteria, be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.

The likelihood is proportional to the pdf of the underlying sample that
enter the analysis sample, as expressed by
\begin{align}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}_S(O | M, M_I) & \propto p(O, \text{in analysis}|M, M_I)\\
&  \propto \prod_{\alpha \in transients} p(O^\alpha, \text{in analysis}|M, M_I),
\end{align}
where the second line holds because the observations and stochastic decisions
for sample selection are expected to be independent from transient to transient.

For a single supernova labeled by $\alpha$,   the probability of the analysis population is related
to the underlying population by
\begin{align*}
p_S(O^\alpha|M, M_I) &\equiv p(O^\alpha, \text{in analysis}|M, M_I)\\
& = P(disc=True, anal=True, sobs=True, type=True, T^\alpha=SNIa|O^\alpha, M, M_I)p(O^\alpha|M, M_I)\\
& = P(disc=True, anal=True, sobs=True, type=True|T^\alpha=SNIa, O^\alpha, M, M_I)\\
& \times P(T^\alpha=SNIa| O^\alpha, M, M_I)p(O^\alpha|M, M_I)\\
&= P(disc=True|T^\alpha=SNIa,c^\alpha, M, M_I)
P(anal=True|T^\alpha=SNIa,c^\alpha, M, M_I) \\
&  \times P(sobs=True|T^\alpha=SNIa,c^\alpha, M, M_I)
 P(type=True|T^\alpha=SNIa, s^\alpha, M, M_I)\\
&  \times
P(T^\alpha=SNIa| M, M_I)p(c^\alpha|M, M_I)p(s^\alpha|M, M_I)p(z^\alpha|M, M_I).
\end{align*}
where we anticipate that the stochastic processes that enter each sample-selection step
are independent of each other, and that the measurement uncertainties are independent.
We find that $p_S$ is the pdf of the underlying
distribution $p(O^\alpha|M, M_I)$ multiplied by the efficiency of a realized target to
enter the analysis sample.

The likelihood can be expressed as
\begin{align}
\mathcal{L}_S(O_o | M, M_I) & = \frac{\prod_{\alpha \in transients}p_S(O_o^\alpha|M, M_I)}{\int  \prod_{\alpha \in transients} p_S(O^\alpha |M, M_I)dO}.
\label{likelihood:eqn}
\end{align}
A frequentist analysis can proceed with this likelihood.
Alternatively, the Bayesian posterior is specified by parameter priors as 
\begin{align*}
p(M, M_i | O_o) & \propto \mathcal{L}_S(O_o| M, M_I) p(M_I|M) p(M)\\
 & \propto \mathcal{L}_S(O_o| M, L, \tau) p(L| \tau, M) p(\tau|M) p(M).
\end{align*}
Once all the $p$'s and $P$'s are specified we are ready to analyze!

@@

\subsection{$p(M_I|M)$: Probabilities of the Intermediary Parameters.}
\begin{itemize}
\item Type probabilities are relevant in the photometric analysis.
The probability an object is SN~Ia is
\begin{align}
P_{SNIa} &= \frac{r_{SNIa}}{r_{SNIa}+r_{SNII}} \nonumber \\
P_{SNII}&=1-P_{SNIa}.
\label{prob:eqn}
\end{align}

@@

There are parameters that specify the sub-type.
\begin{align}
t_0 & \sim \mathcal{U}\\
x_{0, SNIa} & \sim \mathcal{N}(0,R^{x_0}_{SNX})\\
x_{1,SNIa} & \sim \mathcal{N}(x_{1,SNIa}^\star,R^{x_1}_{SNIa})\\
c_{SNIa} & \sim \text{SkewNormal}(c^\star_{SNIa},R^{c}_{SNIa} )
\end{align}

\subsection{$p(O| M, M_I)$: Probability of All Possible Observables}
\begin{itemize}
\item The spectroscopic typing is assumed to be perfect
\begin{equation}
T | \tau \sim \delta (T-\tau).
\label{typepdf:eqn}
\end{equation}

\item The spectroscopic redshift of a spectroscopically typed transient assumed to be perfect
\begin{equation}
z | z_{host} ~ \delta (z-z_{host}).
\end{equation}

@@

\item  The luminosity distance comes from the physics-theory function
\begin{equation}
d_L = d_L(z; \Omega_M, w0),
\end{equation}
the flux for an observation at date $t$ is
\begin{equation}
f = 10^{-0.25 \left(M_{SNIa} - \alpha x_1 + \beta c \right)}\frac{1}{4\pi d_L^2} F_{SALT2}((t-t_0)/(1+z); x_0,  x_1, c, \lambda/(1+z)),
\end{equation}
and expected counts is
\begin{equation}
c_\star = 10^{Z/2.5}f.
\end{equation}
The distributions of counts are 
\begin{align*}
c | c_\star &\sim \mathcal{N}(c_\star, C_c)\\
s | c_\star &\sim \mathcal{N}(c_\star, \sigma_s),
\end{align*}
where $C_c$ and $C_s$ are the measurement covariances for the photometric and spectroscopic observations respectively.

@@

\begin{itemize}
\item $P(T=SNIa| M, M_I) = \delta_{\tau, SNIa}$: The transient is
measured to be SN~Ia if it is SN~Ia.
\item $ P(type=True|T=SNIa, s, M, M_I)$: 
The probability of successful spectroscopic typing needs to be determined.
A simple possibility is that  it only depends on the signal-to-noise of the spectroscopic observation
$\epsilon_T(ston)$.  Then given a  spectroscopic noise of $\sigma_s$
\begin{equation}
P(type=True|  T=SNIa, s, M, M_I) = \epsilon_T(s/\sigma_s).
\end{equation}

\item $P(disc=True|T=SNIa,c, M, M_I)$:
The probability for discovery of a transient in a single image depends on the
signal-to-noise $\epsilon_D(ston)$.  The probability of discovery given all
the images is
\begin{equation}
P(disc=True | T=SNIa,c,  M, M_I)
=  1-\prod_{\beta \in images} \left(1-\epsilon_D(c_\beta/N_\beta)\right).
\end{equation}

\item $P(anal=True|T=SNIa,c, M, M_I)$:
As an illustrative example we write out the probability of a analysis-sample selection requiring two photometric points above
a signal-to-noise threshold $STON_T$: other sample selection cases are similar but with different combinatorics
\begin{align*}
P(anal=True  | T=SNIa, c,  M, M_I)= &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\\
&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T)),
 \end{align*}
 where $\mathcal{H}$ is the Heaviside step function.

\item $P(sobs=True|T=SNIa,c, M, M_I)$:
The probability a transient has a spectroscopic observation.
Mat Smith and Chris D'Andrea should be able to provide this term.
Until we get this, lets make the simplifying ansatz that
\begin{align*}
P(sobs=True | T=SNIa, c, M, M_I)= & const.
 \end{align*}

@@

>>>>

There are top-level  sets of parameters $M$ that correspond to photometric calibration, cosmology,
host-galaxy properties, and transient properties.
There are two sets of intermediary parameters  $M_I=\{L, \tau\}$. that correspond to luminosity and type respectively.

@@

We are interested in what the model predicts for the pdf of those observables
for the population of transients that enter the analysis.  
The likelihood can be expressed in terms of probabilities of the observables
for all possible transients and observables accommodated by the model.
The analysis population differs from the underlying population in that the transients of the former 
must, for the typed SN~Ia analysis, be discovered, pass photometric selection criteria, be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.
The likelihood of the observations $O_o$ is
\begin{align*}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}_S(O_o | M, M_I) & = p(O_o |\text{in analysis}, M, M_I)\\
& = \frac{ p(O_o, \text{in analysis}|M, M_I)}{p(\text{in analysis}|M, M_I)}\\
&  = \frac{\prod_{\alpha \in transients} p(O_o^\alpha, \text{in analysis}|M, M_I)}{p(\text{in analysis}|M, M_I)},
\end{align*}
where the last line holds because the observations and stochastic decisions
for sample selection are expected to be independent from transient to transient.

For a single supernova labeled by $\alpha$,   the probability of the analysis population is related
to the underlying population by
\begin{align}
p_S(O^\alpha|M, M_I) &\equiv p(O^\alpha, \text{in analysis}|M, M_I) \nonumber \\
& = P(disc=True, anal=True, sobs=True, type=True, T^\alpha=SNIa|O^\alpha, M, M_I)p(O^\alpha|M, M_I) \nonumber \\
& = P(disc=True, anal=True, sobs=True, type=True|T^\alpha=SNIa, O^\alpha, M, M_I) \nonumber \\
& \times P(T^\alpha=SNIa| O^\alpha, M, M_I)p(O^\alpha|M, M_I) \nonumber \\
&= P(disc=True|T^\alpha=SNIa,c^\alpha, M, M_I)
P(anal=True|T^\alpha=SNIa,c^\alpha, M, M_I)  \nonumber \\
&  \times P(sobs=True|T^\alpha=SNIa,c^\alpha, M, M_I)
 P(type=True|T^\alpha=SNIa, s^\alpha, M, M_I) \nonumber \\
&  \times
P(T^\alpha=SNIa| M, M_I)p(c^\alpha|M, M_I)p(s^\alpha|M, M_I)p(z^\alpha|M, M_I).
\label{p_S:eqn}
\end{align}
where we anticipate that the stochastic processes that operate on each sample-selection decision
are independent of each other, and that the measurement uncertainties of type, counts,
spectrum, and redshift are independent.
$p_S$ is equal to the pdf of the underlying
distribution $p(O^\alpha|M, M_I)$ multiplied by the efficiency of a realized target to
enter the analysis sample.

The likelihood is then  re-expressed as
\begin{align}
\mathcal{L}_S(O_o | M, M_I) & = \frac{\prod_{\alpha \in transients}p_S(O_o^\alpha|M, M_I)}{\int  \prod_{\alpha \in transients} p_S(O^\alpha |M, M_I)dO},
\label{likelihood:eqn}
\end{align}
where  the denominator $p(\text{in analysis}|M, M_I)$ is set
by the condition that the integral over the likelihood over all possible
observations must equal to one.
A frequentist analysis can proceed with this likelihood.
Alternatively, the Bayesian posterior is specified by parameter priors as 
\begin{align}
p(M, M_i | O_o) & \propto \mathcal{L}_S(O_o| M, M_I) p(M_I|M) p(M) \nonumber \\
 & \propto \mathcal{L}_S(O_o| M, L, \tau) p(L| \tau, M) p(\tau|M) p(M).
\label{posterior:eqn}
\end{align}
Once all the $p$'s and $P$'s introduced in
Eqns.~\ref{p_S:eqn}, \ref{likelihood:eqn}, and \ref{posterior:eqn}
are specified we are ready to analyze!

@@

\subsection{$p(M_I|M)$: Probabilities of the Intermediary Parameters.}
\begin{itemize}
\item $p(\tau| M, M_I) $: Type probabilities are relevant in the photometric analysis.
The probability an object is SN~Ia is
\begin{align}
P_{SNIa} &= \frac{r_{SNIa}}{r_{SNIa}+r_{SNII}}  \\
P_{SNII}&=1-P_{SNIa}.
\label{prob:eqn}
\end{align}

@@

\item $P(L| \tau, M, M_I) $: 
For the SALT2 model several parameters
must be specified to define the subtype,
including the parameter $x_{0, SNIa}$ that corresponds to $L$.
\begin{align}
t_0 & \sim \mathcal{U}\\
x_{0, SNIa} & \sim \mathcal{N}(0,R^{x_0}_{SNX})\\
x_{1,SNIa} & \sim \mathcal{N}(x_{1,SNIa}^\star,R^{x_1}_{SNIa})\\
c_{SNIa} & \sim \text{SkewNormal}(c^\star_{SNIa},R^{c}_{SNIa} )
\end{align}
\end{itemize}

\subsection{$p(O| M, M_I)$: Probability of All Possible Observables}
\begin{itemize}
\item $P(T=SNIa| M, M_I) $: The spectroscopic typing is assumed to be perfect
\begin{equation}
T | \tau \sim \delta (T-\tau).
\label{typepdf:eqn}
\end{equation}

\item $P(z| M, M_I) $: The spectroscopic redshift of a spectroscopically typed transient assumed to be perfect
\begin{equation}
z | z_{host} ~ \delta (z-z_{host}).
\end{equation}

@@

\item  $P(c| M, M_I)$, $P(s| M, M_I)$:
The luminosity distance comes from the physics-theory function
\begin{equation}
d_L = d_L(z; \Omega_M, w0),
\end{equation}
the flux for an observation at date $t$ is
\begin{equation}
f = 10^{-0.25 \left(M_{SNIa} - \alpha x_1 + \beta c \right)}\frac{1}{4\pi d_L^2} F_{SALT2}((t-t_0)/(1+z); x_0,  x_1, c, \lambda/(1+z)),
\end{equation}
and expected counts is
\begin{equation}
c_\star = 10^{Z/2.5}f.
\end{equation}
The distributions of counts are 
\begin{align}
c | c_\star &\sim \mathcal{N}(c_\star, C_c)\\
s | c_\star &\sim \mathcal{N}(c_\star, \sigma_s),
\end{align}
where $C_c$ and $C_s$ are the measurement covariances for the photometric and spectroscopic observations respectively.

@@

\begin{itemize}
\item $ P(type=True|T=SNIa, s, M, M_I)$: 
The probability of successful spectroscopic typing needs to be determined.
A simple possibility is that  it only depends on the signal-to-noise of the spectroscopic observation
$\epsilon_T(ston)$.  Then given a  spectroscopic noise of $\sigma_s$
\begin{equation}
P(type=True|  T=SNIa, s, M, M_I) = \epsilon_T(s/\sigma_s).
\end{equation}

\item $P(disc=True|T=SNIa,c, M, M_I)$:
The probability for discovery of a transient in a single image depends on the
signal-to-noise $\epsilon_D(ston)$.  The probability of discovery given all
the images is
\begin{equation}
P(disc=True | T=SNIa,c,  M, M_Ii)
=  1-\prod_{\beta \in images} \left(1-\epsilon_D(c_\beta/N_\beta)\right).
\end{equation}

\item $P(anal=True|T=SNIa,c, M, M_I)$:
As an illustrative example we write out the probability of a analysis-sample selection requiring two photometric points above
a signal-to-noise threshold $STON_T$: other sample selection cases are similar but with different combinatorics
\begin{align}
P(anal=True  | T=SNIa, c,  M, M_I)= &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T)) \nonumber \\
&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T)),
 \end{align}
 where $\mathcal{H}$ is the Heaviside step function.

\item $P(sobs=True|T=SNIa,c, M, M_I)$:
The probability a transient has a spectroscopic observation.
Mat Smith and Chris D'Andrea should be able to provide this term.
Until we get this, lets make the simplifying ansatz that
\begin{align}
P(sobs=True | T=SNIa, c, M, M_I)= & const.
 \end{align}

@@

\subsection{$p(\text{in analysis}|M, M_I)$: The Normalization}
